Class differences in social networks: Evidence from a referral experiment

### 2025-08-04 writing/manuscript.tex
v1.1.5

Referral hiring—the formal or informal process where firms ask workers to recommend qualified candidates for job opportunities—is a common labor market practice that makes differences in social capital evident.\footnote{Referrals solve some frictions in the search and matching process and benefit both job-seekers and employers. As a consequence, referral candidates get hired more often, have lower turnover, and earn higher wages \citep{brown2016informal,dustmann2016referral,friebel2023employee}.} Since referrals originate from the networks of referrers, the composition of referrer networks becomes a crucial channel that propagates inequality. Similar individuals across socio-demographic characteristics form connections at higher rates \citep{mcpherson2001birds}, making across-SES (low-to-high) connections less likely than same-SES connections \citep{chetty2022na}. Referrals will thus reflect similarities in socio-demographic characteristics present in networks even in the absence of biases in the referral procedure—that is, even when referring randomly from one's network according to some productivity criteria.

@@

Referral hiring in the labor market can range from firm-level formal referral programs asking employees to bring candidates to simply passing on job opportunities between network members \citep{topa2019social}. Since our participants are students at the university and refer based on exam scores, we abstract away from formal referral programs with defined job openings. Our setting instead resembles situations where contacts share opportunities with each other without requiring the referred candidate to take any action and without revealing the referrer's identity. This eliminates reputational concerns since there is no hiring employer. It also establishes a lower bound on the expected reciprocity for the referrer when combined with pay-for-performance incentives \citep{witte2021workers,bandiera2009social}. At the same time, referring based on university entry exam scores is still an objective, widely accepted measure of ability. We show evidence that referrers in our setting not only possess accurate information about these signals but can also screen more productive individuals from their university network.

@@

The key advantage of this approach is that by conditioning on each referrer's enrollment network, we eliminate selection bias from program choice and other factors that determine who appears in each person's choice set. The identifying variation comes from within-network differences in referral decisions, holding constant the pool of available candidates.

We estimate separate models for each referrer SES group to estimate aggregate SES biases across socioeconomic groups.

>>>>

Research on economic connectivity has focused on the relationship between individual choice and chance in meeting high-SES individuals. The prevailing hypothesis emerging from the seminal work of \citet{chetty2022n} is that increasing exposure to high-SES individuals under favorable inter-group contact conditions will lead low-SES individuals to connect with them at higher rates. Universities, in this regard, represent a particularly promising setting since they attract higher-than-population shares of high-SES students, and create more opportunities for cross-SES connections. However, credible evidence on biases in individual choices to connect across SES groups remains limited. One important reason for this gap is the challenge of creating controlled environments that isolate SES biases while accounting for natural variations in social network compositions.

We overcome this challenge through a lab-in-the-field experiment at a Colombian university. We recruited 734 undergraduate students to make incentivized referrals among peers they encountered during their coursework. Referrals were made for the math and critical reading areas of the national university entry exam, and to incentivize performance-based referral selection, participants earned payments up to \$60 per referral based on their nominee's percentile ranking at the university. This setup provided an objective performance benchmark for referrals while SES biases in referral selection could still play a role.

Referrals originated from each participant's unique course enrollment network that we constructed using extensive administrative data. The enrollment network covered each course the referrer had taken with all other undergraduate students at the university (more than 4500 individuals). It allowed us to observe both characteristics of every potential referral candidate, and the number of courses taken together as a measure of the intensity of interaction between the referrer and the candidate. Referrals from the enrollment networks enabled us to separate network composition (i.e., chance of meeting during coursework and frequency of contact) from SES biases in referral selection (i.e., individual choice in picking a referral). By doing so, we were able to control for naturally varying network compositions with referral candidates at the individual level, and could identify group-level SES biases in referral selection that go beyond mere opportunities to interact at the university.

We randomized participants into two conditions. In the \textbf{Baseline} condition participants made referrals with performance-based incentives only, where their earnings depended on the actual performance of their referrals. In the \textbf{Bonus} condition, participants made referrals with performance-based incentives and an additional fixed bonus (\$25) going to their referral of choice. We designed the \textbf{Bonus} condition to make SES biases in referral selection even more salient. The fixed bonus provided more incentives to refer someone who performs less well, but is better connected with the referrer in terms of contact intensity which could correlate with SES.

We find that referrals consistently go to higher-performing peers with strong contact intensity (14 vs. 4 courses taken together), regardless of the incentive conditions and the exam area. Pooling across conditions and exam areas, we find that SES bias in referrals is primarily driven by low-SES participants exhibiting in-group preferences: Controlling for network composition, low-SES referrers are 45\% more likely to refer other low-SES peers  and 44\% less likely to refer high-SES relative to middle-SES peers. In contrast, middle- and high-SES referrers show no in-group biases and also no biases against low-SES peers.

With 93\% of referrals within the same academic program to peers with high numbers of courses taken together, we find that chances to interact during coursework explains most of the observed referral patterns in terms of SES: at the contact intensity where referrals occur (median 12 courses together), low-SES networks contain 44.5\% low-SES peers versus 35\% university-wide (increase of 27\%), while high-SES networks contain only 15.7\% low-SES peers (decrease of 55\%). Even in the absence of a bias against low-SES, this intense network segregation makes it clear that in the range where high-SES referrers consider candidates, low-SES are practically not even in the choice sets for consideration.

Looking for potential mechanisms driving the segregation in enrollment networks, we identify program selection as key. Program fees at the partner university are fixed on a cost basis, and with less than 5\% of undergraduates qualifying for scholarships. The consequence of these institutional policies is that SES groups end up sorting into programs on the basis of their costs, where some programs cost up to six times more on a yearly basis. To sum, even though low-SES are exposed to higher-than-population shares of high-SES students, and high-SES are not biased toward other SES groups, meaningful interaction opportunities are limited due to institutional and organizational factors.

% One potential concern in interpreting our results is that referrals in our scenario have no real consequences in terms of job opportunities. While it is true that the referrals made in our setting do not result in job opportunities, our findings replicate the typical results from the experimental literature studying referrals where performance-based incentives brings in qualified candidates \citep{beaman_who_2012,pallais_why_2016,witte2021workers}, with potential biases in referral selection in play \citep{beaman_job_2018,hederos2025gender}. Moreoever, referral hiring captures a wide range of practices spanning from formal referral programs within hiring firms to informal sharing of job opportunities between two candidates \citep{topa2019social}. The lower-stakes referral environment with objective performance benchmarks could actually provide a lower-bound in SES biases in referral selection.

% Another potential concern is about the nature of the enrollment network. One could question what exactly the enrollment network captures and whether it overlaps with the actual university social network of students. These are tough questions with no way to test in our date. What we can do is comapre our approach with other methods. While there are other ways to elicit networks by simply asking participants for a list of friends, these may be prone to biases in recall, could bias referral decisions, and are limited in the size of networks they elicit. Another solution to this problem could involve collecting participant data from social media platforms (e.g., Facebook) but this would also at best remain a proxy for actual social interactions. Our approach takes advantage of extensive administrative data which is not affected by biases in recall, and arguably overlaps with university social networks that could be collected through social media platforms.

% A final related concern is our measure of contact intensity. One could question whether the number of courses taken together captures the true intensity of social interactions. Our data provides key insights into what this measure actually captures: In a robustness exercise, we find that beyond identifying peers in the same academic programs, the number of courses taken is a significant predictor of who gets a referral. This indactes that the number of courses taken together is a meaningful indicator of social contact between the referrer and the referral.

Our findings contribute to understanding SES biases in referral selection with some important scope conditions. First, while our referrals have no direct job consequences, the performance-based incentive structure replicates typical findings from earler referral experiments \citep{beaman_who_2012,pallais_why_2016}, and the lower-stakes environment may actually provide a lower bound on SES biases compared to high-stakes hiring contexts.

Second, our enrollment networks capture classroom-based interactions rather than broader social networks. This approach offers advantages over self-reported friendship networks, which suffer from recall bias and size limitations, or social media networks, which may not reflect meaningful interactions. The administrative data reveals that course-taking intensity predicts referral selection even beyond program affiliation, suggesting it captures meaningful social contact.

Finally, our setting examines SES bias within a single institution where cross-SES contact is possible. The generalizability to contexts with less SES diversity or different institutional structures remains an open question for future research.

—is a common labor market practice that makes differences in social capital evident.\footnote{Referrals solve some frictions in the search and matching process and benefit both job-seekers and employers. As a consequence, referral candidates get hired more often, have lower turnover, and earn higher wages \citep{brown2016informal,dustmann2016referral,friebel2023employee}.} Since referrals originate from the networks of referrers, the composition of referrer networks becomes a crucial channel that propagates inequality. Similar individuals across socio-demographic characteristics form connections at higher rates \citep{mcpherson2001birds}, making across-SES (low-to-high) connections less likely than same-SES connections \citep{chetty2022na}. Referrals will thus reflect similarities in socio-demographic characteristics present in networks even in the absence of biases in the referral procedure—that is, even when referring randomly from one's network according to some productivity criteria.

@@

@@

The key advantage of this approach is that by conditioning on each referrer's enrollment network, we eliminate selection bias from program choice and other factors that determine who appears in each person's choice set. The identifying variation comes from within-network differences in referral decisions, holding constant the pool of available candidates. We estimate separate models for each referrer SES group to estimate aggregate SES biases across socioeconomic groups.
